---
title: Open Science in Business Economics
author: | 
  | Joachim Gassen 
  | Humboldt-Universität zu Berlin
  | TRR 266 "Accounting for Transparency"
date:  March 17, 2020
  
output: 
  beamer_presentation

header-includes:
- \usepackage{booktabs}
- \usepackage{graphicx}
- \usepackage{xcolor}
- \usepackage{array}
- \usepackage{longtable}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{adjustbox}
- \setbeamertemplate{itemize subitem}{-}
- \titlegraphic{\includegraphics[width=6cm]{media/trr266_logo_white_background.png}}
- \definecolor{trr266petrol}{RGB}{27,138,143}
- \definecolor{trr266blue}{RGB}{110,202,226}
- \definecolor{trr266yellow}{RGB}{255,180,59}
- \definecolor{trr266red}{RGB}{148,70,100}
- \newcommand{\code}{\texttt}
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, table.align = "center",  message = FALSE, error = FALSE, warning = FALSE, clean = FALSE)
library(knitr)
library(kableExtra)
library(tufte)
library(tidyverse)
library(ExPanDaR)
library(rdfanalysis)

opts_knit$set(fig.pos = 'h')
source("../code/utils.R")

# The following allows one to use Stata in RMarkdown 
# Nice but not open science ;-)
# original version
# devtools::install_github("hemken/Statamarkdown")
# Fork that fixed Mac bug non finding the Stata executable
# devtools::install_github("remlapmot/Statamarkdown",  ref = "macos-bug-fixes")
# library(Statamarkdown)

```

```{r prepare_data, include=FALSE}
read_csv("../data/case_study_data.csv", col_types = cols()) %>%
  mutate_at(c("country", "year"), as.factor) %>%
  mutate_at(c("life_expectancy", "gdp_capita",
              "mn_yrs_school", "unemployment"), list(ln = log)) -> smp_all

smp_no_na <- smp_all %>% na.omit

var_def <- read_csv("../data/case_study_data_def.csv", col_types = cols())

for (var in c("life_expectancy", "mn_yrs_school", 
              "mn_yrs_school", "unemployment")) {
  var_def <- rbind(var_def,
                   list(paste0(var, "_ln"), 
                        paste("Log-transformed version of", var), "numeric", 1))
}
```

## Agenda

- What is open science and why should we care?
- Introducing our case: The Preston curve
- Sorry but your code does not run: Research containers 
- Making data FAIR: Issues in economics
- Working around unFAIR data: The 'ExPanDaR' package
- Increase research transparency: The 'rdfanalysis' package


# What is open science and why should we care?

## Open Science

Open Science is the practice of science in such a way that others can collaborate and contribute, where research data, lab notes and other research processes are freely available, under terms that enable reuse, redistribution and reproduction of the research and its underlying data and methods.

`r quote_footer('--- FOSTER, https://www.fosteropenscience.eu/')`


## Too good to be true?

\begin{center}
\includegraphics[height=0.8\textheight]{media/fanelli_2010_PLoS_one.jpg} \\
Fanelli: “Positive” Results Increase Down the Hierarchy of the Sciences (PloS, 2010)
\end{center}


## Data and code repositories are on the rise \dots

\begin{center}
\includegraphics[width=0.8\textwidth]{media/gertler_galiani_romero_nature_fig1_2018.jpg} \\
Gertler, Galiani and Romero (Nature, 2018)
\end{center}


## \dots but yet fail to guarantee reproducible results

\begin{center}
\includegraphics[height=0.8\textheight]{media/gertler_galiani_romero_nature_fig2_2018.jpg} \\
Gertler, Galiani and Romero (Nature, 2018)
\end{center}


## But: The narrative of a "research crisis" ... 

\begin{center}
\includegraphics{media/fanelli_pnas_2018.jpg} \\
Fanelli (PNAS, 2018)
\end{center}


## ... might be exaggerated

The new “science is in crisis” narrative is not only empirically unsupported, but also quite obviously counterproductive. Instead of inspiring younger generations to do more and better science, it might foster in them cynicism and indifference. Instead of inviting greater respect for and investment in research, it risks discrediting the value of evidence and feeding antiscientific agendas.

...

Therefore, contemporary science could be more accurately portrayed as facing “new opportunities and challenges” or even a “revolution”. Efforts to promote transparency and reproducibility would find complete justification in such a narrative of transformation and empowerment, a narrative that is not only more compelling and inspiring than that of a crisis, but also better supported by evidence.

`r quote_footer('--- Fanelli (PNAS, 2018)')`


## Instead: Open Collaboration is a success story!

- Wikipedia
- GNU/Linux
- Git(hub)
- R/RStudio
- Python
- ...

# Introducing our case: \break The Preston curve

## Life expectancy is associated with GDP per capita

\begin{center}
\includegraphics[height=0.7\textheight]{media/preston_1975_p235.png}
Preston (1975): The Changing Relation between Mortality and level
of Economic Development, Population Studies (29): 235.
\end{center}


## Descriptive Statistics

```{r tab_descriptives_sample, echo = FALSE, results="asis", out.height="\\textheight"}
df <- smp_no_na %>% 
  select(life_expectancy, gdp_capita, mn_yrs_school, unemployment) %>%
  mutate(gdp_capita = gdp_capita/1000)

t <- prepare_descriptive_table(df)

desc_rnames <- c(
  "\\textit{Life expectancy}", 
  "\\textit{GDP per capita}", 
  "\\textit{Years of schooling}", 
  "\\textit{Unemplyoment rate}"
)
rownames(t$df) <- desc_rnames
names(t$df)[c(5,7)] <- c("25 \\%", "75 \\%")
kable(t$df, digits = c(0, 1, 1, 1, 1, 1, 1, 1), format = "latex",
      format.args = list(decimal.mark = ".", big.mark = ",", 
                         scientific=FALSE),
      booktabs = TRUE, escape = FALSE, linesep = "") -> kab_latex

lat_tab <- unlist(strsplit(kab_latex, "\n"))
lat_tab[3] <- "\\\\[-1.8ex]\\hline \\hline \\\\[-1.8ex]"
lat_tab[5] <- "\\hline\\\\[-1.8ex]"
lat_tab[length(lat_tab) - 1] <- "\\\\[-1.8ex]\\hline \\hline \\\\[-1.8ex]"

latex_tab <- c("\\begin{threeparttable}",
               lat_tab[2:length(lat_tab)],
               "\\begin{tablenotes}[flushleft]",
               "\\setlength{\\labelsep}{0pt}",
               sprintf("\\item Note: The data is obtained from the World Bank and the Wittgenstein Center. The sample covers %d countries and the period %d to %d. \\textit{GDP per captia} are in constant 2010 thousand U.S. dollars.", length(unique(smp_no_na$country)),
                       min(as.numeric(as.character(smp_no_na$year))), 
                       max(as.numeric(as.character(smp_no_na$year)))),
               "\\end{tablenotes}",
               "\\end{threeparttable}")

cat("\\centering \\adjustbox{max height=\\dimexpr\\textheight-6cm\\relax, max width=\\textwidth}{")
cat(paste(latex_tab, collapse = "\n"))  
cat("}")
```


## Regression Results

```{r estimate_models, include=FALSE}
res <-  prepare_regression_table(
  smp_no_na,
  dvs = rep("life_expectancy", 4),
  idvs = list(
    "gdp_capita_ln", 
    c("gdp_capita_ln", "mn_yrs_school_ln", "unemployment_ln"),
    c("gdp_capita_ln", "mn_yrs_school_ln", "unemployment_ln"),
    c("gdp_capita_ln", "mn_yrs_school_ln", "unemployment_ln")
  ),
  feffects = list("", "", "year", c("country", "year")),
  clusters = list("", "", "year", c("country", "year")),
  format = "latex"
)
min_est = min(confint(res$models[[1]]$model)[2, 1],
              confint(res$models[[2]]$model)[2, 1],
              confint(res$models[[3]]$model)[1, 1],
              confint(res$models[[4]]$model)[1, 1])
              
max_est = max(confint(res$models[[1]]$model)[2, 2],
              confint(res$models[[2]]$model)[2, 2],
              confint(res$models[[3]]$model)[1, 2],
              confint(res$models[[4]]$model)[1, 2])
```

```{r tab_regression_results, echo = FALSE, results='asis'}

vars_out <- c("gdp\\_capita\\_ln", "mn\\_yrs\\_school\\_ln", 
              "unemployment\\_ln", "country", "year")
vars_in <- c("ln(\\textit{GDP per capita})", "ln(\\textit{Years of schooling})",
             "ln(\\textit{Unemployment rate})", "Country", "Year")
tb <- res$table

for (i in 1:length(vars_out)) {
  tb <- str_replace_all(tb, fixed(vars_out[i]), vars_in[i])
}


tb[10] <- " & \\multicolumn{4}{c}{Dependent variable: \\textit{Life expectancy}} \\\\ "  
latex_tab <- c("\\small",
               "\\begin{threeparttable}",
               "\\setlength{\\tabcolsep}{2pt}",
               "\\begin{tabular}{lcccccccc}",
               tb[8:11], tb[13:27], tb[29:35],
               "\\end{tabular}",
               "\\begin{tablenotes}[flushleft]",
               "\\setlength{\\labelsep}{0pt}",
               "\\item Note: The dependent variable is the average life expectancy at birth in years. OLS coefficients are reported together with standard errors in parentheses. $^{*}$/$^{**}$/$^{***}$ indicate two-sided significance levels of 10/5/1 \\%, respectively.",
               "\\end{tablenotes}",
               "\\end{threeparttable}")

cat("\\centering \\adjustbox{max height=\\dimexpr\\textheight-6cm\\relax, max width=\\textwidth}{")
cat(paste(latex_tab, collapse = "\n"))  
cat("}")
```


# Sorry but your code does not run: \break Research containers 

## What is the problem?

- All code is based on a development environment (e.g., your computer)
- Not everybody has access to your development environment (matter-of-factly
most likely nobody has)
- Designing projects so that they can be easily ported across environments is hard.
Too hard for most of us.
- The easier way: Develop your project and its code in a standardized environment. 
Ship that environment with your code.


## Meet Docker

\begin{center}
\includegraphics[width=0.3\textwidth]{media/docker_logo_400x400.png} 
\end{center}
* Take a look at this file: https://github.com/joachim-gassen/opsci_talk/blob/master/docker/Dockerfile


# Making data FAIR: Issues in economics

## What makes data FAIR?

- Data is easily \textcolor{trr266petrol}{findable} both by experts in the field 
and by researchers from other domains

- Data is \textcolor{trr266petrol}{accessible} for interested researchers 
meaning that they can access, understand and use the data

- Data is \textcolor{trr266petrol}{inter-operable} so that it can be merged with 
other data sources

- Data is \textcolor{trr266petrol}{reusable} so that it can be used for other
research projects  besides the research project for which they were initially
collected


## Why is it hard to establish FAIR data in economics?

- Many data are commercialized, especially in the area of capital markets
- Survey data often contains personal information 
- Experimental data is collected based on informed consent from participants
that did not include later sharing of the data
- Data is stored on author's web pages or similar, with very little meta data,
leading to data that only experts in the field are able to find and understand
- Lack of clear identifiers for main units of analyses make it inherently hard 
to merge data from different sources
- Lack of information on data generating processes reduces re-usability


# Working around unFAIR data: \break The 'ExPanDaR' package

## The idea

- When data cannot be shared, you can still provide an interface to the data
so that others can explore it
- That way you enable users to assess the robustness of empirical evidence without providing them with access to the underlying data
- Besides that, the 'ExPanDaR' package also provides a toolbox for researchers to explore data 
on the fly, allowing them to download R notebook code that reflects their analysis

## Meet the ExPanD app

\begin{center}
\includegraphics[height=0.7\textheight]{media/expand_preston.png} \\
https://jgassen.shinyapps.io/preston
\end{center}

# Increase research transparency: \break The 'rdfanalysis' package

## Empirical research requires making choices

\begin{center}
\includegraphics[height=0.7\textheight]{media/leamer_aer_1983_37f.png} \\
Leamer (AER 1983: 37f.)
\end{center}


## Slightly more systematic ...

\begin{center}
\includegraphics[height=0.8\textheight]{media/wicherts_et_al_2016_table_1_p3.png} \\
Wicherts et al. (Frontiers in Psychology 2016: 3)
\end{center}


## Objective of the `rdfanalysis` package

Development of a statistical computing framework for a "multiverse analysis" (Steeger et al., Persp on Psych Sci 2016) that

- supports research design development separate from data,
- facilitates a priori power analysis,
- promotes unit testing,
- generates well documented and easily portable code,
- makes researcher degrees of freedom (Simmons et al., Psyc Science 2011) explicit in code, and
- allows for rigorous robustness checking by exhausting all these degrees of freedom algorithmically. 

Note: Code based walk-through on what follows: https://joachim-gassen.github.io/rdfanalysis/articles/analyzing_rdf.html


## Workflow

1. Develop your \textcolor{black}{\code{design}} by specifying
the \textcolor{trr266yellow}{\code{result}} that you want to estimate and the 
required \textcolor{trr266petrol}{\code{steps}} to do that.
2. The package generates a code skeleton based on your 
\textcolor{black}{\code{design}}.
3. Implement each \textcolor{trr266petrol}{\code{step}} starting from this
skeleton. Assign positive weights to all design choices that you consider to
be supported best by theory ex ante.
4. Use package functions to document and test your code for internal consistency.
5. Optional: Run code on simulated data to test its computational correctness and 
assess the power of your \textcolor{black}{\code{design}}.
6. Run \textcolor{black}{\code{design}} with real data and generate a weighted estimate of the \textcolor{trr266yellow}{\code{result}}.
7. Exhaust \textcolor{black}{\code{design}} across all 
\textcolor{trr266red}{\code{protocols}} and use the
visualization tools of the package to assess the robustness of your \textcolor{trr266yellow}{\code{result}} across all researcher degrees of freedom.


## The case steps and their discrete/continuous choices

\begin{center}
\includegraphics[height=0.85\textheight]{media/rdf_design.png}
\end{center}


## Estimate the effect size for the weighted protocols

``` {r exhaust_weighted_ests, include = "FALSE", results = "hide"}
load("../raw_data/case_study_results.Rdata")
```
``` {r table_weighted_results, echo = FALSE, results='asis'}

kable(weighted_ests, digits = 3, booktabs = TRUE, linesep = "",
      format = "latex") -> kab_latex
lat_tab <- unlist(strsplit(kab_latex, "\n"))

cat("\\centering \\adjustbox{max height=\\dimexpr\\textheight-6cm\\relax, max width=\\textwidth}{")
cat(paste(lat_tab, collapse = "\n"))  
cat("}")

```

And the weighted average estimate is:

``` {r estimate_weighted_effect, echo = FALSE}
weighted_eff <- calculate_weighted_estimate(weighted_ests, "est", "lb", "ub")
kable(as.data.frame(weighted_eff), digits = 3, booktabs = TRUE, linesep = "")
```


## Some but not all of the researcher degrees of freedom

```{r SepcCurve, cache = TRUE, echo = FALSE, fig.align="center", out.width="\\textwidth"}
df <- ests %>%
  filter(na.omit == "no",
         outlier_cutoff == 0 | outlier_cutoff == 0.01 | outlier_cutoff == 0.05,
         cluster == "ctryyear") %>%
  select(-na.omit, -cluster)
attr(df, "choices") <- 1:5
plot_rdf_spec_curve(df, "est", "lb", "ub", 
                    est_label = expression(Delta*" Life expectancy (yrs)"), pt_size = 0.5,
                    choice_ind_point = FALSE) 
```


## And an interactive variant of it

\bcenter

https://jgassen.shinyapps.io/shiny_rdf_spec_curve/

\ecenter


## For further info and code

* Code for this talk: \textcolor{beamer@blendedblue}{\url{https://github.com/joachim-gassen/opsci_talk}}

* 'ExPanDaR' package: \textcolor{beamer@blendedblue}{\url{https://joachim-gassen.github.io/ExPanDaR}}

* 'rdfanalysis' package: \textcolor{beamer@blendedblue}{\url{https://joachim-gassen.github.io/rdfanalysis}}

* Online course for open science methods: \textcolor{beamer@blendedblue}{\url{https://github.com/joachim-gassen/sposm}}

* Blog on R and Open Science 
\textcolor{beamer@blendedblue}{\url{https://joachim-gassen.github.io}}

* Twitter: \textcolor{beamer@blendedblue}{\url{https://twitter.com/JoachimGassen}}
